---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Vicky Fong"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
install.packages("agricolae")
library(tidyverse); library(lubridate); library(here); library(agricolae); library(dplyr)
getwd()

ntl <- read.csv(
  here('Data','Raw','NTL-LTER_Lake_ChemistryPhysics_Raw.csv')
  )

ntl$year4 <- as.factor(ntl$year4)
ntl$sampledate <- mdy(ntl$sampledate)

#2
my_theme <-  theme(
    plot.title = element_text(face = "bold", size = 12),
    panel.background = element_rect(fill = "white", colour = NA), 
    panel.border = element_rect(fill = NA, colour="grey50"), 
    panel.grid.major = element_line(colour = "black", size = 0.01),
    panel.grid.minor =  element_line(colour = "black", size = 0.01),
    axis.text = element_text(size = 10),
    axis.ticks = element_blank(),
  )

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July will not change with depth across all lakes. 
Ha: Mean lake temperature recorded during July will decrease as depth increaeses across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ntl <- ntl %>%
  mutate(month = month(sampledate)) %>%
  filter(month == 7) %>%
  select(lakename,year4,daynum,depth,temperature_C) %>%
  na.omit()

#5
ntl %>%
  ggplot(aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  labs(x="Depth (m)", y="Temperature (C)", 
       title="Temperature and Depth of Lakes in the Month of July") +
  ylim(0,35) +
  my_theme
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure supports our alternative hypothesis that temperature would decrease with increasing depth. The inverse relationship between temperature and depth is most evident from 0 to 5 meters depth, and the strong negative slope suggessts that the greatest temperature change occurs in the first 5 meters of the lake depth. The range of temperatures between 5 to 10 meters is smaller, suggesting less temperture changes occur in this depth range. From 10 to 15 meters, most of the temperature points are around 5 degrees C.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
ntl.regression <- lm(data = ntl, temperature_C ~ depth)
summary(ntl.regression)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The R-squared value of 0.7387 means that 73.9% of the variabilitty is explained by changes in depth. This finding is based ono 9726 degrees of freedonm. Since the P-value is less than 0.05, it suggests a statistically significant relationship between temperature and depth of lakes in the month of July. The coefficient of the slope is -1.94621, which means tempeature is predicted to decrease by 1.95 degrees C for every 1 meter increase in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
ntl.AIC <- lm(data = ntl, temperature_C ~ depth + year4 + daynum)
step(ntl.AIC) #include all variables

#10
summary(ntl.AIC)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC recommended including all available variables (depth, year and day number) to predict temperature in the multiple regression as the step function only provided one option. Multiple R-squared is 0.7604, suggesting this model explains 76.0% of the observed vaariance. This is a slight improvement from the model with depth as the only variable at 73.9%.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
ntl$lakename <- as.factor(ntl$lakename)
ntl.anova <- aov(data = ntl, temperature_C ~ lakename)
summary(ntl.anova)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, the p-value of the anova is less than 0.05, suggesting that there is a significant difference in mean temperature among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ntl %>%
  ggplot(aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha=0.5) +
  geom_smooth(method=lm, se=FALSE) +
  labs(x="Depth (m)", y="Temperature (C)", color="Lake",
       title="Temperature and Depth of Lakes in the Month of July") +
  ylim(0,35) +
  my_theme
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(ntl.anova)
ntl.groups <- HSD.test(ntl.anova, "lakename", group = TRUE)
ntl.groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The Tukey's HSD test found that Ward Lake and Peter Lake had the same mean temperature, statistically speaking, as Peter Lake. There were no lakes that had a statistically distinct mean temperature. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can use two-sample t-test to see if Peter Lake and Paul Lake have statistically distinct mean temperatures.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
crampton.ward <- ntl %>% 
  filter(lakename == c("Crampton Lake", "Ward Lake"))
two.t <- t.test(crampton.ward$temperature_C ~ crampton.ward$lakename)
two.t
```

>Answer: The two-sample t-test had a p-value of 0.3263, which is greater than 0.05, thus we are unable to reject our null hypothesis. This suggests that there is no statistical difference between the July temperatures in Crampton Lake and Ward Lake. The mean temperatures of the lakes are not equal - 15.4 for Crampton Lake and 14.3 for Ward Lake - but not different enough to be statistically significant. This matches the results from the Tukey HSD as Crampton Lake and Ward Lake both belonged in group "b", which means that mean temperatures are statistically similar enough for them to be grouped together for analysis.
